---
title: "Module 14 - Feature-Level Analyses"
output: html_document
---

In this R Markdown script, we will load in our taxonomy phyloseq object from the Moving Pictures Tutorial and we will run:

-   Core microbiome

-   Indicator species analysis

-   Differential abundance

The last few modules have showcased a lot of wrangling from Module 10 and elsewhere. From this point onward, the scripts will stripped to their essentials - it will be up to you to apply code from the previous modules if you want to explore or wrangle the data further!

```{r Load data and libraries}
library(tidyverse)
library(phyloseq)

# Load object
ps = readRDS('../Datasets/phyloseq_taxonomy.rds')
```

# Core microbiome

I am listing the libraries as I go so that you can see where each one is used.
However, best practice is to load all libraries at the start of the script.

**Consider:** How could you modify the following code so that you only write out each command after transform() once?

```{r}
library(microbiome) # Same package as the transform() command

# Rarefy if desired and convert to relative abundance, use tax_glom (optional)
psrare = rarefy_even_depth(ps, sample.size = 2513)
ps_rare_relab = transform(psrare, 'compositional')
ps_rare_relab_genus = tax_glom(ps_rare_relab, 'Genus')

# Subset your phyloseq object
site.gut = subset_samples(ps_rare_relab_genus, body.site == 'gut')
site.tongue = subset_samples(ps_rare_relab_genus, body.site == 'tongue')
site.leftpalm = subset_samples(ps_rare_relab_genus, body.site == 'left palm')
site.rightpalm = subset_samples(ps_rare_relab_genus, body.site == 'right palm')

# Find core members
ASVs_gut = core_members(site.gut, detection=0.001, prevalence = 0.3)
ASVs_tongue = core_members(site.tongue, detection=0.001, prevalence = 0.3)
ASVs_leftpalm = core_members(site.leftpalm, detection=0.001, prevalence = 0.3)
ASVs_rightpalm = core_members(site.rightpalm, detection=0.001, prevalence = 0.3)

library(ggVennDiagram)
ggVennDiagram(list(ASVs_gut, ASVs_tongue, ASVs_leftpalm, ASVs_rightpalm),
      			  set_size = 6,
              category.names = c('gut','tongue','left palm','right palm') )
```

# Indicator species analysis

```{r statistics}
library(indicspecies)

# Aggregate ASVs to higher taxonomic level (optional), 
# convert phyloseq to relative abundance, 
# apply abundance filter.

ps_phylum = tax_glom(ps,'Phylum')
ps_relab = transform(ps_phylum, 'compositional')
ps_filt = filter_taxa(ps_relab, function(x) mean(x) > 0.001, TRUE)
otu_table = data.frame(otu_table(ps_filt))

set.seed(421)
indval = multipatt( t(otu_table), 
            cluster= ps_filt@sam_data$body.site,
 		  control = how(nperm = 999) )

summary(indval, indvalcomp = TRUE)

# Extract into table
indval_table = as.data.frame(indval$sign)
```

```{r plot}
# Plot the two taxa that are indicators for the gut
phyla_to_plot = indval_table %>% 
  filter(s.gut==1 & s.tongue==0 & `s.left palm`==0 & `s.right palm`==0 ) %>% 
  rownames()

df_of_taxa = prune_taxa(phyla_to_plot, ps_filt) %>% psmelt()

df_of_taxa %>% 
  ggplot(aes(body.site,Abundance,fill=body.site)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(height=0, width=0.2) +
  facet_wrap(~Phylum, ncol = 4, scales = 'free')
```

# Differential abundance with Maaslin2

```{r statistics}
# if(!requireNamespace("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")
# BiocManager::install("Maaslin2")
library(Maaslin2)

# Tax_glom if desired. DON'T rarefy or use a compositional transformation!
ps_glom = tax_glom(ps, 'Genus')

set.seed(421)
out = Maaslin2(
  # Your count table and metadata, extracted into data frames
  input_data = data.frame(ps_glom@otu_table), 
  input_metadata = data.frame(ps_glom@sam_data), 
  # Maaslin2 forces you to specify an output directory, where it automatically saves some outputs. I usually just delete these results because I don't like the formatting.
  output = 'to_delete', 
  # List out your explanatory variable(s) - ex. c('var1','var2')
  fixed_effects = c('body.site'),
  # Optional: specify a reference group for your categorical variable(s). 
  # This will be the group that all other groups are compared to. 
  # Ex. if you have a variable called 'Treatment' with two groups 'Control' and 'Intervention', you would set reference = 'Treatment,Control' to assess Intervention relative to Control.
  reference = 'body.site,gut',
  # TSS is Total Sum Scaled, AKA relative abundance.
  normalization = 'TSS',
  # Arcsine transformation is recommended for relative abundance data.
  transform = 'AST',
  # Set abundance, prevalence, Padj value thresholds
  min_abundance = 0.001, # 0.001 means 0.1% relative abundance
  min_prevalence = 0.1, # 0.1 means 10% of samples
  max_significance = 0.05, # Padj threshold
  # These plots increase run time and are usually poorly formatted, so you might as well stop them from being generated
  plot_heatmap = FALSE,
  plot_scatter = FALSE
)

statistical_table = out$results

writexl::write_xlsx(statistical_table,'../Results/Tables/Module 14 - Maaslin2 Results.xlsx')
```

```{r Plot}
# Filter stats to only include taxa that are differentially abundant between the gut and tongue
taxa_to_plot = statistical_table %>% 
  filter(qval < 0.05)

# BASIC PLOT ~~~~~~~~~~~~~

# This will plot the EFFECT SIZE (coefficient) for the significant taxa.

# This is NOT the same thing as log 2 fold changes, which is what you should actually plot for your final project. However, log2FC are not provided by Maaslin2.
# I am leaving the details of how to calculate log 2 fold changes from the relative abundance data as an exercise for you - you can do it manually (see Assignment 6 answer key) or you can use the fix_lfc() function that I wrote in ggpicrust2_calculate_log2fc.R.

# Don't forget to:
#    A) plot the genera names instead of the representative ASV names
#    B) make it publication-ready (ex. sorting the bars from largest to smallest)
taxa_to_plot %>%
  ggplot(aes(feature,coef)) +
  geom_col() +
  coord_flip()
```